---
title: "13: High Frequency Data"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Determine stormflow and baseflow from high frequency flow data
2. Use a hysteresis plot to understand watershed dynamics
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

High frequency data is usually defined as frequencies significantly lower than daily (e.g. 5-minute, 15-minute, 1 hr etc). What types of hydrological and biological processes happen on this timescale that we might want to investigate?

> Discharge (storm events) Oxygen content changes throughout the day; sediment transport; temp changes; contaminant transport;


## Session Set Up

```{r, message = FALSE}
getwd()

packages <- c(
  "dataRetrieval", "tidyverse", "cowplot", "EcoHydRology", "xts", "dygraphs"
  )
invisible(lapply(packages, library, character.only = TRUE)) 

theme_set(theme_classic(base_size = 12))
```

## Pull Data

Still from USGS, but this time readNWISuv = instantaneous values
```{r download data}

dischargedatTFC2018 <- readNWISuv(
  site = "02097280", #Third Fork Creek at Woodcroft Parkway Near Blands
  parameterCd = "00060",
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns() #automatically renamed values based on the NWIS database definitions (they already know)
#the date column is a POSIXct type of variable, not date. Because it is also storing time. 
#the universal time zone is from Greenwich. 


dischargedatEC2018 <- readNWISuv(
  site = "0208675010", #Ellerbe Creek at Club Boulevard at Durham, NC
  parameterCd = "00060", 
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()



plot_grid(
  ggplot(dischargedatTFC2018, aes(x = dateTime, y = Flow_Inst)) +
    geom_line() + 
    scale_y_log10() +
    ggtitle("Third Fork Creek"),
  ggplot(dischargedatEC2018, aes(x = dateTime, y = Flow_Inst)) + 
    geom_line() + 
    scale_y_log10() +
    ggtitle("Ellerbe Creek"),
  ncol = 1
)
          

#we are separating baseflow and quickflow.
#baseflow follows all the troughs of the storm peaks
#quickflow is all the peaks
```

The shape and size of river hydrographs following rain events can tell us how water moves through a watershed. Once rain falls, how does water get to a stream? What would the streamflow do?

> overland flow, groundwater, fall in river channel


_"Hydrograph separation is one of the most desperate analysis techniques in use in hydrology" - Hewlett & Hibbert 1967_


```{r baseflowseparation}
TFCbaseflow <- BaseflowSeparation(
  dischargedatTFC2018$Flow_Inst, 
  filter_parameter = 0.925, #default parameter
  passes = 3 #default parameter
  )
#gives us two columns: bt(baseflow time series) and qft (quickflow time series)

TFC2018 <- cbind(dischargedatTFC2018, TFCbaseflow) #adding back in dates

ggplot(TFC2018, aes(x = dateTime, y = Flow_Inst)) + 
  geom_line() +
  # scale_y_log10() +
  geom_line(mapping = aes(x = dateTime, y = bt), color = "darkorange4") +
  geom_line(mapping = aes(x = dateTime, y = qft), color = "steelblue4")

dygraph(
  cbind(
    Flow = with(TFC2018, xts(Flow_Inst, order.by = dateTime)), 
    Baseflow = with(TFC2018, xts(bt, order.by = dateTime)), 
    Quickflow = with(TFC2018, xts(qft, order.by = dateTime))
    )
  ) %>% 
  dyRangeSelector()
#baseflow: comes slower, from groundwater that percolates and then eventually makes its way through soil into the stream
#quickflow is the first water in the storm - overland flow

Export <- TFC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_), #what is time difference in between each of these time stamps? dateTime is in POSIXct format. as.numeric changes it to the number of seconds between two time stamps. added NA value added at end because a difference gives us one less point.
         baseflowexport = bt * timestep, #volume = rate (bt) x time
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T), #sum of all volume over the year
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)

#export shows the total amount of water exported from the stream
  
```

What percentage of water is exported from this watershed as baseflow? What percentage as quickflow?

> 51.9% exported as baseflow and 48.1% exported as quickflow

```{r}
Export$BaseflowExport_cf/Export$TotalExport_cf

1-(Export$BaseflowExport_cf/Export$TotalExport_cf)

```


Repeat this analysis for Ellerbe Creek. How do the percentages compare?


```{r}
ECbaseflow <- BaseflowSeparation(
  dischargedatEC2018$Flow_Inst, 
  filter_parameter = 0.925, #default parameter
  passes = 3 #default parameter
  )


EC2018 <- cbind(dischargedatEC2018, ECbaseflow) #adding back in dates

Export <- EC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_),
         baseflowexport = bt * timestep, #volume = rate (bt) x time
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T), #sum of all volume over the year
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)
```

```{r}
Export$BaseflowExport_cf/Export$TotalExport_cf

1-(Export$BaseflowExport_cf/Export$TotalExport_cf)

```

> 43.4% as baseflow, and 56.6% quickflow

What does this mean physically about the watershed?

> Ellerbe creek has more overland flow - closer to center of city, TFC has more trees, more open area because it has more baseflow.

## Hydrologic Flashiness

A "flashy" stream increases in flow faster following a precipitation event. 

Refering to our previous work with baseflow separation, would you expect flashiness to increase or decrease with urbanization?

> urbanization increases flashiness

What do you think "urban stream syndrome" refers to?

> Very flashy, tend to be deeper(insized), water quality parameters that are characteristic of urban streams

The Richards-Baker Index is an index of how flashy a stream is. It is calculated as

$$RBI = (\sum_{j = 1}^n |Q_j - Q_{j-1}| / \sum_{j = 1}^n Q_j) / Catchment Area$$

summing up total differences in flow between time stamps, divided by total flow, divided by catchment area
the smaller the catchment, the shorter amount of time it takes to get to the stream

```{r Flashiness}
TFCsitedata <- readNWISsite(site = "02097280")
TFC.catchment.size <- TFCsitedata$drain_area_va #square miles = 14.79

RBI_TFC2018 <- (
  sum(abs(diff(dischargedatTFC2018$Flow_Inst))) / 
    sum(dischargedatTFC2018$Flow_Inst[-1])
  ) / TFC.catchment.size

ECsitedata <- readNWISsite(site = "0208675010")
EC.catchment.size <- ECsitedata$drain_area_va #square miles

RBI_EC2018 <- (
  sum(abs(diff(dischargedatEC2018$Flow_Inst))) / 
    sum(dischargedatEC2018$Flow_Inst[-1])
  ) / EC.catchment.size

RBI_TFC2018
RBI_EC2018
```

Which stream is flashier?

> Ellerbe Creek is flashier because it has a higher RBI value


Now, do with another site of your choosing:
```{r}
COdischarge <- readNWISuv(
  site= "09163500", 
  parameterCd = "00060",
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()

CObaseflow <- BaseflowSeparation(
  COdischarge$Flow_Inst, 
  filter_parameter = 0.925, #default parameter
  passes = 3 #default parameter
  )


CO2018 <- cbind(COdischarge, CObaseflow) #adding back in dates

ExportCO <- CO2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_),
         baseflowexport = bt * timestep, #volume = rate (bt) x time
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T), #sum of all volume over the year
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)
view(ExportCO)

ExportCO$BaseflowExport_cf/ExportCO$TotalExport_cf

1-(ExportCO$BaseflowExport_cf/ExportCO$TotalExport_cf)

```


## Chemical Flashiness

Chemical concentrations in rivers can be either highly variable with discharge, or "chemostatic", which means the concentration barely changes with discharge. 

```{r}
CCdat <- readNWISuv(
  site = "02249500", # CRANE CREEK AT MELBOURNE, FL
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2015-01-1",
  endDate = "2016-01-01"
  ) %>%
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(CCdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()


Potdat <- readNWISuv(
  site = "01646500", # POTOMAC RIVER NEAR WASH, DC LITTLE FALLS PUMP STA
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2016-01-1",
  endDate = "2017-01-01"
  ) %>% 
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(Potdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

Which of these rivers is more chemostatic?

> Crane Creek

